Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-25 paper/paper.tex
Edited this intro. I couldn't leave it in an non-optimal form.

This paper presents the first Coq formalisation to our knowledge of a recursion scheme, called
the \emph{hylomorphism}, along with its algebraic laws allowing for the
mechanisation of all recognised (terminating) recursive algorithms. The key
contribution of this paper is that this formalisation is fully axiom-free
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted OCaml code for the programs formalised in our
framework is efficient, resembles code that a human programmer would write, and
contains no occurrences of \ocaml{Obj.magic}.  We also present a machine-checked
proof of the well-known short-cut fusion optimisation.
\end{abstract}

@@

The main approach for implementing non-structural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}.  Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on an
\emph{accesibility} predicate on the input, which is a proof that there are no
infinitely decreasing chains.}. A benefit of this approach is that, extracting
verified nonstructural recursive functions to OCaml will erase the termination
proofs, and so produce code that will be closer to what a programmer may have
written directly in OCaml. However, reasoning about program equivalences
requires dealing with such termination proofs, and it is common practice to use
custom reduction lemmas that will be used extensively when proving properties
about nonstructural recursive functions.

\emph{Structured recursion schemes} have been successfully employed in
functional programming to structure recursive programs, for example, quicksort
and mergesort are both instances of a divide-and-conquer algorithm which in
terms of recursion schemes can be can be formalised as an
hylomorphism~\cite{MeijerFP91, HuIT96}. Furthermore, it has been shown by Hinze
et al.~\cite{HinzeWG15} that hylomorphisms provide the basic building block of
every recursion scheme. In particular, any complex recursion scheme can be
transmogrified down into an hylomorphisms by means of an adjunction.
Furthermore, hylomorphism laws can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to semi-automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.
% Recursion schemes have successfully been applied to functional programming to
% structure recursive programs, however, because of the presence of
% non-termination in languages like (e.g.) Haskell, they cannot guarantee
% recursive definitions well-defined. % This is due to the possibility of applying a
% %
% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
% % use of data functors.

However, when used in the context of languages with general recursion like
(e.g.) Haskell, recursion schemes cannot ensure termination, but while
formalising recursion schemes in a type theory does provide stronger termination
guarantees, to the best of our knowledge, not many attempts have been made at
mechanising structured recursion schemes.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is problematic since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation to our knowledge of
\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).

>>>>

To the best of our knowledge, this paper presents the first Coq formalisation of
a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
laws allowing for the mechanisation of all recognised (terminating) recursive
algorithms. The key contribution of this paper is that this formalisation is
fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
exemplify the framework by formalising a series of algorithms based on different
recursive paradigms such as divide-and conquer, dynamic programming, and mutual
recursion and demonstrate that the extracted OCaml code for the programs
formalised in our framework is efficient, resembles code that a human programmer
would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
a machine-checked proof of the well-known short-cut fusion optimisation.
\end{abstract}

@@

The main approach for implementing non-structural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are accompanied by
\emph{termination proofs}. Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on an
\emph{accesibility} predicate on the input, which is a proof that there are no
infinitely decreasing chains.}.
A benefit of this strategy is that when extracting verified non-structural
recursive functions to OCaml, the additional proofs are discarded, resulting in
code that closely resembles what a programmer might have written directly in
OCaml. However, reasoning about program equivalences needs  such termination
proofs to be kept around, leading to the common practice of employing additional
reduction lemmas extensively.

% A benefit of this approach is that, extracting
% verified nonstructural recursive functions to OCaml will erase the additional
% proofs, thus producing code that will be closer to what a programmer may have
% written directly in OCaml. However, reasoning about program equivalences
% requires dealing with such termination proofs, and it is common practice to use
% custom reduction lemmas that will be used extensively when proving properties
% about nonstructural recursive functions.

\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
for structuring recursive definitions that has found a lot of success in
functional programming. The idea is that, rather than writing a self-refential
definition, we use a function library, the recursion scheme, which takes as
input the non-recursive parts of the algorithm and returns the desired recursive
definition which is guaranteed to terminate by the properties of the recursion
scheme itself. More complex recursion schemes do exist but it has been
proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
This result, thus, yields a general framework to reason about the algebraic properties of these
algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.

% A more complex recursion scheme is the hylomorphism~\cite{MeijerFP91, HuIT96} which has a divide-and-conquer
% structure. It has been shown by Hinze et
% al.~\cite{HinzeWG15} that any complex recursion scheme can be transmogrified
% into an hylomorphisms by means of an adjunction. Furthermore, hylomorphism laws
% can capture a number of useful equivalences, ranging from common optimisations
% such as \emph{short-cut fusion}~\cite{TakanoM95}, to semi-automatic
% parallelisations~\cite{Gibbons96:Third, farmsCastro}.

% Recursion schemes have successfully been applied to functional programming to
% structure recursive programs, however, because of the presence of
% non-termination in languages like (e.g.) Haskell, they cannot guarantee
% recursive definitions well-defined. % This is due to the possibility of applying a
% %
% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
% % use of data functors.

However, when used in the context of languages with general recursion like
(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is because the
input information itself could be non-terminating. On the other hand, while
formalising recursion schemes in a type theory does provide stronger termination
guarantees, to the best of our knowledge, not many attempts have been made at
mechanising them.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
program reasoning, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
